---
title: "final_proj"
author: "christian otto"
date: "5/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.See what distribution of gdp_per_capita / suicides_percentage is across countries.

```{r 1}
library(tidyverse)
suicide_df <- read_csv("suicide.csv")
head(suicide_df)
```

```{r 2}
suicide_df <- suicide_df %>% rename('gdp_per_capita' = 'gdp_per_capita ($)' , 'suicide/100k' = 'suicides/100k pop') %>%
  select("country","year","sex","age","suicide/100k","gdp_per_capita", "suicides_no", "population") 
head(suicide_df)
```



```{r 3}
suicide_by_year_df <- suicide_df %>% 
  group_by(country, year) %>%
  summarize(suicide_no = sum(suicides_no), population = sum(population), gdp_per_capita = sum(gdp_per_capita)/n()) %>%
  ungroup()
head(suicide_by_year_df)
```

```{r 4}
suicide_by_year_df <- suicide_by_year_df %>% 
  mutate(suicide_percentage = suicide_no / population * 100, suicide_percent_per_100gdp = suicide_percentage/gdp_per_capita * 100) 
head(suicide_by_year_df)
```


Just by the spread, you'd think we'd have a roughly inverse relationship between the percent of a population committing suicide and the gdp_per_capita, but the linear regression line reveals that there might be a slightly positive linear relationship between the two or maybe no relationship at all.
```{r 5}
suicide_by_year_df %>% ggplot(aes(x = gdp_per_capita, y = suicide_percentage)) + geom_point() + geom_smooth(method=lm)
```
Here's a better look at the distribution of gdp_per_capita we saw in the last plot. 
```{r 6}
suicide_by_year_df %>% ggplot(aes(x = gdp_per_capita)) + geom_histogram()
```

Another look
```{r 7}
suicide_by_year_df %>% ggplot(aes(y = gdp_per_capita)) + geom_boxplot()
```

```{r 8}
suicide_by_year_df %>% filter(year %in% c("1990","1995", "2000", "2010","2015")) %>%
  ggplot(aes(x=gdp_per_capita, y=suicide_percentage)) +
  facet_grid("year") +
  geom_point() + 
  geom_smooth(method=lm)
```

Now gdp_per_capita across time
```{r 9}
suicide_by_year_df %>% ggplot(aes(x = year, y = gdp_per_capita)) + geom_point() + geom_smooth(method=lm)
```
And suicide rates across time. This contradicts the claim that there's a relationship between gdp per capita and suicide rates, since, on average, gdp steadily climbed while suicide percentage stayed constant.
```{r 10}
suicide_by_year_df %>% ggplot(aes(x = year, y = suicide_percentage)) + geom_point() + geom_smooth(method=lm)
```


```{r 11}
suicide_by_year_df %>% ggplot(aes(x = year, y = suicide_percent_per_100gdp)) + geom_point() + geom_smooth(method=lm)
```


